---
title: "Final Project"
author: "Kauanoe Greene"
date: "2024-12-04"
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.path = "../Output/")
```

# Libraries  

```{r}

# visuals
library(tidyverse)
library(dplyr)
library(tidytext)
library(here)
library(gganimate)

# models (stats)
library(lmerTest)
library (RLRsim)
library(Matrix)
# library(lme4)
library(lattice)
library(car)
library(nlme)
```

# Data  

```{r}

# Data uploading
functional.phase.data <- read_csv(here("Final_Project", "Data", "functional.phase.data.csv"))
glimpse(functional.phase.data)
view(functional.phase.data)
```

# Statistic Modeling: Chlorophyll Content Data  

- PT.I: Primary interaction treatment  
- PT.II: Reduced models  
- PT.III: Comparison models  

## PT.I: Full Model 
- main interaction  
- includes all main factors and interactions  
- F-statistic for a fixed effects mixed model  
- results table: drought column   

```{r}

# drought column result: (F-stat) 3.9797, (p-val) 0.07637

chl.full.A = lmer(Chlorophyll ~ Treatment + 
                    (1|ID) + 
                    (1|Population) + 
                    (1|Week) + 
                    (1|Population:Treatment) +  
                    (1|Treatment:Week), 
                  data = functional.phase.data, 
                  na.action = na.omit)

Anova(chl.full.A, test.statistic="F", type=3)
```


```{r}

# Normality check: untransformed data  
# Checking normality of residuals in untransformed data
# W = 0.98162, p-value = 1.319e-10

shapiro.test(resid(chl.full.A)) 
```


```{r}

# untransformed data: QQPlot  
qqnorm(resid(chl.full.A))
```


```{r}

# log-transformed data
# this will be used in the next steps when we look at it as log-transformed data.
log.chl <- log(functional.phase.data$Chlorophyll)
glimpse(log.chl)
```


```{r}

# drought column result: (F-stat) 0.5701, (p-val) 0.4732

chl.full.B = lmer(log.chl ~ Treatment + 
                    (1|ID) + 
                    (1|Population) + 
                    (1|Week) + 
                    (1|Population:Treatment) +  
                    (1|Treatment:Week), 
                  data = functional.phase.data, 
                  na.action = na.omit)

Anova(chl.full.B, test.statistic="F", type=3)
```


```{r}

# transformed data: QQPlot 
qqnorm(resid(chl.full.B))
```


## Reduced Models  

- work bottom to top approach  
- drop the interaction  
- keep main factors  
- always keep “Treatment” and “(1|ID)” in all models  
- our goal is to find out if any of these factors improve model fit →   
- does it account for more of the variance in our response trait?  
- if it doesn’t, the chi square value will NOT be significant.  
- the factor we drop from the reduced model is the factor of focus. We want to know if adding this factor accounts for MORE variance than the treatment alone. A.K.A. we are investigating the effect of the factor we drop.  

```{r}

# 1: (Population:Treatment): chisq 6.5799

chl.red.1 = lmer(Chlorophyll ~ Treatment + 
                   (1|ID) + 
                   (1|Population) + 
                   (1|Week) + 
                   (1|Population:Treatment), 
                 data = functional.phase.data, 
                 na.action = na.omit)
Anova(chl.red.1)
```


```{r}

# 2: (Treatment:Time): chisq 4.1081

chl.red.2 = lmer(Chlorophyll ~ Treatment + 
                   (1|ID) + 
                   (1|Population) + 
                   (1|Week) + 
                   (1|Treatment:Week), 
                 data = functional.phase.data,  
                 na.action = na.omit)
Anova(chl.red.2)
```


```{r}

# 3: no interactions: chisq 6.58

chl.red.3 = lmer(Chlorophyll ~ Treatment + 
                     (1|ID) + 
                     (1|Population) + 
                     (1|Week), 
                   data = functional.phase.data, 
                   na.action = na.omit)
Anova(chl.red.3)
```


```{r}

# 4: effect of week: chisq 6.0626

chl.red.4 = lmer(Chlorophyll ~ Treatment + 
                     (1|ID) + 
                     (1|Population), 
                   data = functional.phase.data, 
                   na.action = na.omit)
Anova(chl.red.4)
```


```{r}

# 5: effect of population: chisq 5.4025

chl.red.5 = lmer(Chlorophyll ~ Treatment + 
                         (1|ID) + 
                         (1|Week), 
                       data = functional.phase.data, 
                       na.action = na.omit)
Anova(chl.red.5)
```


## Comparison Models  
- the question: does adding the interaction/factors improve the model fit?
- comp.model.1 = anova(full.model, red.model.1) # test treatment x time interaction  
- comp.model.2 = anova(full.model, red.model.2) # test treatment x population interaction  
- comp.model.3 = anova(red.model.3, red.model.4) # test effect of week  
- comp.model.4= anova(red.model.3, red.model.5) # test effect of population   

```{r}

# 1: Test (treatment:time) interaction: chisq 2.8486
anova(chl.full.A, chl.red.1)
```


```{r}

# 2A: Test (treatment:population) interaction: chisq 0
anova(chl.full.A, chl.red.2)
```

```{r}
# 2B: Test (treatment:population) interaction: chisq 2.8486
anova(chl.red.1, chl.red.2)
```


```{r}

# 3: Test the effect of week: chisq 1.226
anova(chl.red.3, chl.red.4)
```


```{r}

# 4: Test the effect of population: chisq 26.032
anova(chl.red.3, chl.red.5)
```

# Plot  

```{r}
# Data uploading
functional.phase.data <- read_csv(here("Final_Project", "Data", "functional.phase.data.csv"))
glimpse(functional.phase.data)
```


```{r}

# group control and treatment
# group by phase
# summarise mean

glimpse(functional.phase.data)

mean_data <- functional.phase.data %>% 
  select(Week, Population, Treatment, Chlorophyll, Conductance, ETRmax, Phase) %>% 
  group_by(Week, Population, Treatment, Phase) %>% 
  summarize(mean_chl = mean(Chlorophyll, na.rm = TRUE), 
            mean_cond = mean(Conductance, na.rm = TRUE), 
            mean_etrm = mean(ETRmax, na.rm = TRUE))

glimpse(mean_data)

# save csv file to data folder
write.csv(mean_data, here("Final_Project", "Data", "mean_data.csv"))
```


```{r}
# Chlorophyll content over time

chlplot <- mean_data %>% # datasheet
  ggplot(aes(x = Week, # x-axis
             y = mean_chl, # y-axis
             color = Treatment)) + # colors
  geom_point() +  # data points
  geom_line() +  # plot
  labs(subtitle = "Detecting trait plasticity of chlorophyll content in response to drought stress", # plot subtitle
       caption = "Data sourced from: Greene 2023", # plot caption
       x = "Time (Weeks)", # x-axis label
       y = "Mean Chlorophyll Content") + # y-axis label
  ggtitle("Effects of Drought on Leaf Chlorophyll Content in 'A'ali'i") + # plot title
  facet_wrap(~Population) + # create panels for each population!
  scale_color_manual(breaks= c("C", "PD"), labels = c("Control Group", "Pulse Drought Group"), values = c("darkgreen", "brown")) + # rename legend variables
  theme(plot.title = element_text(face = "bold", color = "black", hjust = 0.5), # bold title
        axis.text.x = element_text(size = 8), axis.title.x = element_text(size = 10), # adjust x-axis labels
        axis.text.y = element_text(size = 8), axis.title.y = element_text(size = 10),   # adjust y-axis labels
        legend.position = "top", 
        plot.subtitle = element_text(size = 10, hjust = 0.5), 
        legend.title = element_blank())

  # transition_states(Week)

# save plot to my output folder
ggsave(here("Final_Project", "Output", "chlplot.png")) 

# view plot
chlplot
```

```{r}
# group by phases

phase_data <- functional.phase.data %>% 
  select(Week, Population, Treatment, Chlorophyll, Conductance, ETRmax, Phase) %>% 
  group_by(Population, Treatment, Phase) %>% 
  summarize(mean_chl = mean(Chlorophyll, na.rm = TRUE), 
            mean_cond = mean(Conductance, na.rm = TRUE), 
            mean_etrm = mean(ETRmax, na.rm = TRUE))

glimpse(phase_data)
```


```{r}

# plot recovery phases

chlphase_plot <- phase_data %>% # datasheet
  ggplot(aes(x = Phase, # x-axis
             y = mean_chl, # y-axis
             color = Treatment)) + # colors
  geom_point() +  # data points # plot
  geom_line(aes(group = Treatment)) +
  labs(subtitle = "Investigating the priming effects of drought treatment on rates of chlorophyll content throughout recovery phases", # plot subtitle
       caption = "Data sourced from: Greene 2023", # plot caption
       x = "Recovery Phase", # x-axis label
       y = "Mean Chlorophyll Content") + # y-axis label
  ggtitle("Drought Priming Effect on Leaf Chlorophyll Content in 'A'ali'i") + # plot title
  facet_wrap(~Population) + # create panels for each population!
  scale_color_manual(breaks= c("C", "PD"), labels = c("Control Group", "Pulse Drought Group"), values = c("darkgreen", "brown")) + # rename legend variables
  scale_x_discrete(labels = c('Treatment', 'Pre', 'Mid', 'Post')) +
theme(plot.title = element_text(face = "bold", color = "black", hjust = 0.5), # bold title
        axis.text.x = element_text(size = 8), axis.title.x = element_text(size = 10), # adjust x-axis labels
        axis.text.y = element_text(size = 8), axis.title.y = element_text(size = 10),   # adjust y-axis labels
        legend.position = "top", 
        plot.subtitle = element_text(size = 10, hjust = 0.5), 
        legend.title = element_blank())

# save plot to my output folder
ggsave(here("Final_Project", "Output", "chlphase_plot.png")) 

# view plot
chlphase_plot

```

